---
title: "HYPOTHESIS TESTING"
subtitle: "Data versus the status quo"
author: "Austin Hart"
institute: "American University"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: [default, rladies, rladies-fonts]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
  message=FALSE, warning=FALSE, eval = TRUE, echo = FALSE, 
  fig.align = 'center', dev='svg'
)
```

```{r results='hide'}
library(tidyverse)
library(magrittr)
library(kableExtra)

setwd("~/HypothesisTesting")
df = read_csv("hurricanes.csv")
```

# Topics covered

- Logic of hypothesis testing  

- Meaning of a $p$-value  

- Three implementations
  - $Z$-test for sample mean
  - $t$-test for difference of means
  - $F$-test for difference of group means  
  
- Jung et al (2014). "Female-named hurricanes" 


---
class: inverse, middle, right

# HYPOTHESIS TESTING
### Logic and Process


---
# Hypothesis testing

### Statistical hypothesis testing 

> A method of inference that provides a basis for rejecting a hypothesis based on the plausibility of the data. 
>
> Typically for evaluating "significance" of an association observed in sample data.


---
# Testing procedure

Given a sample statistic from your data:

- Formalize competing hypotheses
  - Alternative, $H_A$: your argument
  - Null, $H_0$: opposite of $H_A$; "status quo" expectation  
  
- Calculate prob of the data given the null, $p = Pr(Sample~Stat | H_0)$  

- Reject/Fail to reject null, typically when $p\leq0.05$  


---
# Setting up hypotheses

- Alternative hypothesis, $H_A$
  - *Your* expectation
  - Empirical implication of your theory

- Null hypothesis, $H_0$
  - Logical opposite of $H_A$
  - Typically the status quo

- TIPS
  - ALWAYS frame hypotheses in terms of parameters
  - It's all about the null. 
  
  
---
# To reject or not to reject?

> On what grounds can I reject the null hypothesis?

- Default position: status quo wins
  - fail to reject the null
  - Stick with conventional wisdom
  
- Reject the null
  - *iff* sample stat is implausible given $H_0$
  - typically $p \leq 0.05$


---
# How is this possible?

> Hypothesis testing is possible because of what we know about the sampling distribution of different sample statistics.

- Pop parameters give shape to the sampling distribution

- Null an argument about parameters *AND* sampling dist

- Allows us to gauge probability of a sample stat


---
class: inverse, middle, right

# THE ALL MIGHTY p-VALUE
### Meaning and use

---
# The meaning of a p-value

> If the conventional wisdom is true, what share of samples reveal a stat or association like yours? 


$$
p = Pr(Sample~Stat | H_0)
$$

A p-value is the probability of observing a sample statistic under the assumption that the null hypothesis is true. 

- Start with your test statistic, e.g. $\hat{p}$

- Measures distance to the hypothesized parameter $\pi_0$

- Standardize by number of standard errors $$ Z = \frac{\hat{p} - \pi_0}{\sigma_p} $$
  
  
---
# p-value as criterion

- Science abhors Type 1 error (throwing away good status quo)

- Reject null only for a massive disconnect b/w conventional wisdom and available evidence

- Gold standard of 0.05
  - Reject null when $p \leq 0.05$
  - Fail to reject when $p > 0.05$


---
class: inverse, middle, right

# THREE HYPOTHESIS TESTS


---
# Population proportions (Z-test)

> Experts think that about 5% of U.S. adults are vegetarian. In a recent survey of 300 adults, you find that 9% identify as vegetarian. Is this significantly larger than the expert assessment?

1. State hypotheses
2. Calculate $p$-value based on a $Z$-test
3. Reject/fail to reject the null


---
# Assessing differences of means (t-test)

> Jung et al. (2014) argue that female-named hurricanes are 
> deadlier than male-named hurricanes. Using the "hurricanes.csv" 
> data, describe the relationship and then evaluate their claim.


---
# Presenting results (t)

The table below explores the relationship between hurricane-name gender and deaths based on data from 94 hurricanes that made landfall in the US from 1950 to 2012. Death totals ranged from a minimum of 0 to a maximum of 1,833 deaths (Katrina, 2005). Female-named hurricanes caused 57 deaths on average. Consistent with Jung et al.'s (2014) argument, male-named hurricanes saw an average of only 15 deaths. However, the difference of means is not statistically significant (t = 1.47, p = 0.07, 1-tailed) and we cannot reject the argument that death totals are unrelated to the storm's name.


```{r ttest}

  df %>%
   group_by(gender) %>%
   summarise(
     n = n(),
     Min = min(deaths),
     Max = max(deaths),
     Median = median(deaths),
     Mean = mean(deaths)
   ) %>%
   kbl(digits = 1, caption = "Deaths by gendered name", type = 'html') %>%
   kable_paper(bootstrap_options = "striped", full_width = F, 
               position = "center")

```

---
# Comparing equality of group means (F-test)

> Use the "hurricanes.csv" data to evaluate the relationship
> between the intensity of the storm at landfall (category) and
> deaths caused. Use an F-test to assess the siginficance.

1. State hypotheses
2. Calculate p-value based on $t$-test
3. Reject/fail to reject the null


---
# Presenting results (F)

The table below summarizes the relationship between the category of storm at landfall and the loss of life. Category 1 storms are the most common and also the least deadly (mean = 13). [more on others] There are only three category 5 storms in the data with more than 700 deaths on average. The differences are statistically significant (F = 10.93, p = 0.001), and we reject the null hypothesis of independence.

```{r Ftest}
df %>%
  group_by(category) %>%
  summarize(
    n = n(),
    Mean = mean(deaths, na.rm = T),
    Median = median(deaths, na.rm = T)
  ) %>%
  na.omit() %>%
  kbl(digits = 1, caption = "Deaths by storm severity", type = 'html') %>%
  kable_paper(bootstrap_options = "striped", full_width = F, 
              position = "center")
```

---
class: inverse

# Wrapping up

- It's you against the status quo (null)  

- Assess the data in light of the null  

- Reject the status quo when what you see no longer makes sense in light of old dogma.
