---
title: "HYPOTHESIS TESTING"
subtitle: "4 tests in action"
author: "Austin Hart"
institute: "American University"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: [default, rladies, rladies-fonts]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
---


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
  message=FALSE, warning=FALSE, eval = TRUE, echo = FALSE, 
  fig.align = 'center', dev = 'svglite'
)
```

```{r results='hide'}
library(tidyverse)
library(kableExtra)
library(janitor)

df = read_csv("hurricanes.csv") %>%
  filter(!is.na(femScale))

load('bechdel.Rdata')
mov = mov %>%
  mutate(Era = if_else(year %in% 1970:1990, "1970-1990","1991-2013"))
```

<style type="text/css">
.remark-slide-content {
    font-size: 25px;
    padding: 1em 4em 1em 4em;
}
</style>

# Hypothesis testing

> A method of inference that provides a basis for rejecting a hypothesis based on the plausibility of the data. 
>
> Typically for evaluating "statistical significance" of an association observed in sample data. 
  
- Four implementations
  - $t$-test for sample means
  - $t$-test for difference of means
  - $t$-test for regression coefficients
  - $\chi^2$-test for tabular analysis
  
- EX: Jung et al (2014). "Female-named hurricanes"  


---
# Terms to know

- Level of significance $\alpha$: probability of Type-I error  

- p-value: probability of a sample statistic given a true null  

- Statistically significant: unlikely to have occurred by chance alone (given a true null)  

- Alternative hypothesis, $H_A$: your expectation about a characteristic of the population  

- Null hypothesis, $H_0$: the opposite of $H_A$. Status quo expectation.


---
# Hypothesis testing procedure

1. Specify hypotheses

2. Set criteria for rejecting the null

3. Compute test statistic  

3. Conclude


---
# Setting up hypotheses

- **Alternative hypothesis,** $H_A$
  - Your expectation of the population/truth
  - Or the empirical implication of your theory

- **Null hypothesis,** $H_0$
  - Logical opposite of $H_A$
  - Status quo expectation: you've found nothing new

- TIPS
  - ALWAYS frame hypotheses in terms of parameters
  - Science is all about the null. 
  
  
---
# Select the right test statistic

> A test statistic is a number that summarizes how far your sample information is from the parameter value in the null hypothesis.  

- Sample proportion: $Z$ statistic
- $t$ statistic for
  - sample mean
  - difference of means
  - OLS regression coefficient
- Tabular analysis: $\chi^2$



---
# To reject or not to reject?

### Set a decision rule for rejecting the null  

- Choose a level of significance, typically $\alpha = 0.05$  

- Find the critical value for your test statistic, e.g., $t_{\alpha/2}$

- Reject null if $t \geq t_{\alpha/2}$



---
# Finding your critical value, 5% significance

.pull-left[
### Upper-tailed test

Reject $H_0$ if $t > 1.645$

```{r sclt1s, fig.width=3.5, fig.height=3, dpi=350}
  ggplot(NULL, aes(c(-3.5,3.5))) +
    geom_area(stat = "function", fun = dnorm, fill = "#00998a", xlim = c(-3, 1.645)) +
    geom_area(stat = "function", fun = dnorm, fill = "grey80", xlim = c(1.645, 3.5)) +
    labs(x = "Z", y = "") +
    scale_y_continuous(breaks = NULL,
                       expand = expansion(mult=c(0,0.05))) +
    scale_x_continuous(breaks = c(0,1.645)) +
    coord_cartesian(xlim = c(-3.5,3.5)) +
    theme_minimal() +
    theme(panel.grid.minor = element_blank())
```

]
.pull-right[
### Two-tailed test

Reject $H_0$ if $\lvert t \rvert > 1.96$


```{r sclt2s, fig.width=3.5, fig.height=3, dpi=350}
  ggplot(NULL, aes(c(-3.5,3.5))) +
    geom_area(stat = "function", fun = dnorm, fill = "#00998a", xlim = c(-1.96, 1.96)) +
    geom_area(stat = "function", fun = dnorm, fill = "grey80", xlim = c(1.96, 3.5)) +
    geom_area(stat='function', fun = dnorm, fill = 'grey80', xlim = c(-3.5,-1.96)) +
    labs(x = "Z", y = "") +
    scale_y_continuous(breaks = NULL,
                       expand = expansion(mult=c(0,0.05))) +
    scale_x_continuous(breaks = c(-1.96,0,1.96)) +
    coord_cartesian(xlim = c(-3.5,3.5)) +
    theme_minimal() +
    theme(panel.grid.minor = element_blank())
```
]


---
class: inverse
# Population mean (t-test)

- I have data for ONE numeric variable.
- I want to test hypotheses about a pop mean. $\mu_y$


---
# Population mean (t-test)

.left-column[
### Setup
]

.right-column[
> On balance, hurricane names are more feminine than masculine. Use Jung et al's ratings to find out.

- State hypotheses
  
- Identify test statistic
  
- Set decision rule
  
- Calculate test statistic
  
- Reject/fail to reject the null

]

---
# Population mean (t-test)

.left-column[
### Setup
### Hypotheses
]

.right-column[
- $H_A:~\mu > 6$
  
- $H_0:~\mu\leq 6$

- Statistics from replication data:  

```{r tpoptab}
df %>%
  summarise(
    N = n(),
    Mean = mean(femScale),
    SD = sd(femScale) 
  ) %>%
  pivot_longer(cols = 1:3, names_to = 'Stat', values_to = 'Fem rating') %>%
  knitr::kable(digits = 1L, caption = 'Sample Statistics')
```

]

---
# Population mean (t-test)

.left-column[

### Setup
### Hypotheses
### t-stats

]

.right-column[

- Appropriate test statistic: 

$$
t=\frac{\bar{y}-\mu_0}{s_y/\sqrt{n}}
$$
  
- Decision rule
  - 5% significance
  - One-sided test
  - Reject null if...
  
]

---
# Population mean (t-test)

.left-column[
### Setup
### Hypotheses
### t-stats
### Reject?
]

.right-column[

- Calculate test statistic, $t$

- Reject or not?

> I reject the null. The mean rating is significantly higher than 6 $(t=2.4)$, meaning that hurricane names are significantly more feminine on balance.

]


---
class: inverse
# Difference of means (t-test)

- I have:
  - Numeric outcome variable
  - Binary exposure variable

- I want to know if the mean outcome differs by group


---
# Difference of means (t-test)

.left-column[
### Setup
]

.right-column[
> Female-named hurricanes deadlier than male-named hurricanes. Describe the relationship, and evaluate the claim.

- State hypotheses
  
- Identify test statistic
  
- Set decision rule
  
- Calculate test statistic
  
- Reject/fail to reject the null

]

---
# Difference of means (t-test)

.left-column[
### Setup
### Hypotheses
]

.right-column[

- $H_A: ~ \mu_{Fem} - \mu_{Male} > 0$
  
- $H_0: ~ \mu_{Fem} - \mu_{Male} \leq 0$
  
- Summary stats from replication data:

```{r diffstats}
df %>%
  group_by(gender) %>%
  summarise(
    n = n(),
    Mean = mean(deaths),
    SD = sd(deaths)
  ) %>%
  pivot_longer(2:4, names_to = ' ') %>% 
  pivot_wider(names_from = gender, values_from = value) %>%
  kable(digits = 1L, caption = 'Deaths by name')
  
```

]

---
# Difference of means (t-test)

.left-column[
### Setup
### Hypotheses
### t-stats
]

.right-column[

- Appropriate test statistic, difference of means

$$
t = \frac{(\bar{y_1}-\bar{y_2})-(\mu_1-\mu_2)}{\sqrt{s_1^2/n_1+s_2^2/n_2}}
$$

- Decision rule
  - For 5% significance
  - One-sided test
  - Reject null if...
]

---
# Difference of means (t-test)

.left-column[
### Setup
### Hypotheses
### t-stats
### Reject?
]

.right-column[


]


---
class: inverse

# Regression

- I have:
  - Numeric outcome variable
  - (typically) Numeric exposure variable
  
- I want to know if the mean outcome is related to level of exposure


---
# Testing regression coefficients

.left-column[

### Setup

]
.right-column[

> Hurricane damage is correlated with the "femininity" of the storm's name.

- State hypotheses
  
- Identify test statistic
  
- Set decision rule
  
- Calculate test statistic
  
- Reject/fail to reject the null

]

---
# Testing regression coefficients

.left-column[

### Setup
### Hypotheses

]
.right-column[

Given linear model: $damage_i = \beta_0 + \beta_1(femScale_i) + e_i$

- $H_0:~ \beta_1 = 0$, $H_A:~ \beta_1 \neq 0$

- Regression estimates from replication data:

```{r ols11}
# Estimate and store
  mfits = list(
    'Deaths' = lm(deaths ~ femScale, data = df),
    'Damage' = lm(damage ~ femScale, data = df)
  )
  cm = c('femScale' = 'Femininity of name',
         '(Intercept)' = 'Constant')
# Present as table
  modelsummary::modelsummary(mfits, fmt = 2, coef_map = cm,
                             title = "Storm 'femininity' and consequences",
                             gof_map = 'nobs', output = 'gt',
                             notes = 'Note: OLS ests with std errors in parentheses.')
  
```

]

---
# Testing regression coefficients

.left-column[

### Setup
### Hypotheses
### t-stats

]

.right-column[

> Hurricane damage is associated with the "femininity" of the storm's name.

- Appropriate test statistic, OLS regression 

$$
t=\frac{b_1 - \beta_1}{SE_b}
$$
  
- Decision rule
  - For 5% significance
  - TWO-sided test
  - Reject null if...
  
]

---
# Testing regression coefficients

.left-column[

### Setup
### Hypotheses
### t-stats
### Reject?

]
.right-column[

```{r ols12}
# Estimate and store
  mfits = list(
    'Deaths' = lm(deaths ~ femScale, data = df),
    'Damage' = lm(damage ~ femScale, data = df)
  )
  cm = c('femScale' = 'Femininity of name',
         '(Intercept)' = 'Constant')
# Present as table
  modelsummary::modelsummary(mfits, fmt = 2, coef_map = cm,
                             title = "Storm 'femininity' and consequences",
                             gof_map = 'nobs', output = 'gt',
                             notes = 'Note: OLS ests with std errors in parentheses.')
  
```

]

---
class: inverse

# Tabular analysis $(\chi^2)$

- I have:
  - Nominal outcome variable
  - Nominal exposure variable
  
- I want to know if the outcome and exposure variable are independent


---
# Tabular analysis

.left-column[
### Setup
]

.right-column[
> Development of female characters differs pre vs post 1990. Bechdel data.

- State hypotheses
  
- Identify test statistic
  
- Set decision rule
  
- Calculate test statistic
  
- Reject/fail to reject the null

]


---
# Tabular analysis

.left-column[
### Setup
### Hypotheses
]


.right-column[

- $H_0:$ Representation is independent of era
  
- $H_A:$ Representation differs by era

- Tabulation in sample:

```{r xtabs}
  xtab =
    mov %>% 
    count(BechdelBinary,Era) %>%  
    na.omit() %>%
    pivot_wider(
      names_from = Era,
      values_from = n, 
      values_fill = 0
    ) %>%
    mutate_if(
      is.integer, list(Percent = ~./sum(.) * 100)
    ) %>% 
    select(BechdelBinary,sort(tidyselect::peek_vars())) %>%
    adorn_totals()

  xtab %>%
    kbl(digits = 1, format = "html", 
        caption = "Bechdel over time", 
        col.names = c(" ","n","%","n","%")) %>%
    kable_paper(bootstrap_options = "striped", full_width = F, position = "center") %>%
    add_header_above(c(" " = 1,"1970-1990"=2,"1991-2013" = 2))
```

]
 
 
---
# Tabular analysis

.left-column[
### Setup
### Hypotheses
### chi-squared stats
]

.right-column[

- Appropriate test statistic

$$
\chi^2_{df} = \sum \frac{(O-E)^2}{E}
$$
  - $O=$ frequency in each cell
    
  - $E=\frac{row.total}{grand.total}*col.total$
    
  - $df=(x.cats-1)*(y.cats-1)$
  
- Decision rule
  - For 5% significance
  - Reject null if...
]
  
  
---
# Tabular analysis

.left-column[
### Setup
### Hypotheses
### chi-squared stats
]

.right-column[

Use data below to caluclate the $\chi^2$ test statistic.

```{r xtabs2}
  xtab =
    mov %>% 
    count(BechdelBinary,Era) %>%  
    na.omit() %>%
    pivot_wider(
      names_from = Era,
      values_from = n, 
      values_fill = 0
    ) %>% 
    select(BechdelBinary,sort(tidyselect::peek_vars())) %>%
    adorn_totals(where = c('row','col'))

  xtab %>%
    kbl(format = "html", 
        caption = "Bechdel over time") %>%
    kable_paper(bootstrap_options = "striped", full_width = F, position = "center")
```
]

